---
title: "Clean International Monetary Fund (IMF) Data"
author: "Yusuf Kiwanuka, Megan Fredrickson, David Hyon, Benjamin Sacks"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
---


```{r load-packages, message=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
```


### Process and Clean World Economic Outlook (WEO) Data

#### Summary  

This code cleans and reshapes data from the **World Economic Outlook (WEO) Database**. It:  

1. Reads the raw Excel file.  
2. Removes unnecessary columns and replaces `"n/a"` or blanks with `NA`.  
3. Converts year columns to numeric for consistency.
4. Reshapes the data to make each economic indicator a separate column.  

#### Source  

The data comes from the **[World Economic Outlook (WEO) Database](https://www.imf.org/en/Publications/WEO/weo-database/2024/October/select-country-group)**, published by the International Monetary Fund (IMF). This resource is a reliable reference for global economic analysis, covering indicators like GDP growth, inflation, and trade balances. It’s widely used by policymakers, researchers, and analysts for understanding trends in the global economy.

```{r clean-weo-data, warning=FALSE}
# Read the Excel file
data <- read_excel("../../data/raw/WEO_Data.xlsx", col_types = "text")

# Replace "n/a" and blank cells with NA, and convert year columns to numeric
data_cleaned <- data %>%
  select(-Units, -Scale) %>%  
  mutate(across(everything(), ~ na_if(., "n/a"))) %>%  # Replace "n/a" with NA
  mutate(across(starts_with("20"), as.numeric))

# Pivot the data to make it wide
data_wide <- data_cleaned %>%
  pivot_longer(cols = starts_with("20"),
               names_to = "Year", 
               values_to = "Value") %>%
  pivot_wider(names_from = `Subject Descriptor`, 
              values_from = "Value")
```

### Add Economic Classification to IMF Data

#### Summary

This code integrates economic classifications into the cleaned World Economic Outlook (WEO) data. It:

1. Reads a CSV file containing country classifications
2. Merges the classification data with the IMF data to include a developing column, coded as binary:
   - `1` for "Emerging and Developing" economies.
   - `0` for "Advanced" economies.
3. Exports the merged dataset to a new CSV file for further analysis.


#### Source

The classification data is derived from the [IMF website](https://www.imf.org/en/Publications/WEO/weo-database/2024/October/groups-and-aggregates) but is not available as a direct download. The lists were manually copied into a spreadsheet and cleaned before use.

```{r add-developing-status}
# Read the country status dataset
country_status <- read.csv("../../data/raw/IMF_country_status.csv")

# Merge IMF data with country status to add "developing" column
imf_data <- data_wide %>%
  left_join(country_status, by = "Country")

# Export the merged data to a CSV file
write.csv(imf_data, file = "../../data/interim/IMF_data.csv", row.names = FALSE)
```

