---
title: "Clean Barro-Lee Educational Attainment Dataset"
author: "Yusuf Kiwanuka, Megan Fredrickson, David Hyon, Benjamin Sacks"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
---


```{r load-packages, message=FALSE}
library(dplyr)
library(tidyr)
```

### Filter Barro-Lee Educational Attainment Data  

#### Summary  

This code focuses on extracting relevant data from the **Barro-Lee Educational Attainment Dataset**. It:  

1. Loads the raw dataset containing global educational attainment metrics.  
2. Filters for specific columns:  
   - `yr_sch`: Average Years of Schooling Attained.  
   - `yr_sch_pri`: Average Years of Primary Schooling Attained.  
   - `yr_sch_sec`: Average Years of Secondary Schooling Attained.  
   - `yr_sch_ter`: Average Years of Tertiary Schooling Attained.  
3. Limits the data to the years 2000–2015.  

#### Source  

The dataset, available [here](https://github.com/barrolee/BarroLeeDataSet/blob/master/BLData/BL_v3_M2564.csv), provides high-quality educational attainment data across 146 countries in 5-year intervals.

```{r filter-barro-lee-data}
# Load the data
data <- read.csv("../../data/raw/BL_v3_MF1564.csv")

# Filter the data
filtered_data <- data %>%
  select(country, year, yr_sch, yr_sch_pri, yr_sch_sec, yr_sch_ter) %>%
  filter(year >= 2000 & year <= 2015)

knitr::kable(head(filtered_data), "simple")
```

### Interpolation Helper Function  

The `interpolate_values` function calculates interpolated and extrapolated values for columns in a dataset:  

1. Expands the year range to include all years from 2000 to 2020.  
2. Interpolates missing values between 2000 and 2015 using linear interpolation.  
3. Extrapolates values beyond 2015 using the slope of the last known segment.  
4. Processes all columns starting with `"yr"`, making it adaptable for various metrics.  

```{r}
# Helper function to calculate interpolated values for a range of years
interpolate_values <- function(df) {
  # Expand the years for interpolation
  full_years <- seq(2000, 2020, by = 1)
  interpolated <- data.frame(year = full_years)
  interpolated$country <- unique(df$country)
    
  # Define columns to interpolate
  yr_columns <- names(df)[startsWith(names(df), "yr")]

  for (col in yr_columns) {
    # Get slopes for interpolation
    slopes <- diff(df[[col]]) / diff(df$year)
    intercepts <- df[[col]][-length(df[[col]])]
    
    interpolated[[col]] <- sapply(interpolated$year, function(x) {
      if (x %in% df$year) {
        df[[col]][df$year == x]
      } else if (x < 2015) {
        segment_index <- max(which(df$year < x))
        intercepts[segment_index] + slopes[segment_index] * (x - df$year[segment_index])
      } else {
        # Extrapolation
        last_slope <- slopes[length(slopes)]
        last_value <- df[[col]][length(df[[col]])]
        last_value + last_slope * (x - max(df$year))
      }
    })
  }
  
  return(interpolated)
}
```

#### Apply Interpolation to the Dataset  

This code applies the `interpolate_values` function to the filtered dataset:  

1. Groups the data by country and splits it into separate data frames.  
2. Applies the interpolation function to each country-specific dataset.  
3. Combines the interpolated results into a single data frame for further analysis.  

```{r}
# Apply to the dataset
full_data <- filtered_data %>%
  group_by(country) %>%
  group_split() %>%
  lapply(interpolate_values) %>%
  bind_rows()

# View the results
knitr::kable(head(full_data), "simple")
```

```{r}
# Export to a CSV file
write.csv(full_data, file = "../../data/interim/BL_data.csv", row.names = FALSE)
```

