---
title: "Query WHO Data"
author: "Yusuf Kiwanuka, Megan Fredrickson, David Hyon, Benjamin Sacks"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_float: true
---


```{r load-packages, message=FALSE}
library(httr)    
library(jsonlite) 
library(tidyr)   
library(dplyr)   
library(purrr)
```


### Define Function: Fetch and Process WHO Indicator Data

#### Description

This code defines the fetch_numeric_data function, which retrieves numeric indicator data from the WHO API for countries starting from the year 2000. It performs the following steps:

#### Process

1. Constructs the API URL using the provided indicator code.
2. Fetches the data via an HTTP GET request, handling errors if the response is invalid.
3. Parses the JSON response and processes the data to include relevant dimensions.
4. Renames key columns (SpatialDim, TimeDim, NumericValue) for clarity and returns a cleaned data frame.

```{r define-fetch-numeric-data}
fetch_numeric_data <- function(indicator_code, column_name, num_dims = 0) {
  num_dims <- as.integer(num_dims)
  if (num_dims < 0 || num_dims > 3) {
    stop("The 'num_dims' parameter must be between 0 and 3.")
  }
  
  # Construct API URL
  url <- paste0(
    "https://ghoapi.azureedge.net/api/",
    indicator_code,
    "?$filter=SpatialDimType%20eq%20'COUNTRY'",
    "%20and%20NumericValue%20ne%20null",
    "%20and%20TimeDim%20ge%202000"
  )
  
  # Fetch data from API
  response <- GET(url)
  if (status_code(response) != 200) {
    stop("Failed to fetch data for indicator:", indicator_code)
  }
  
  # Parse JSON response
  data <- fromJSON(content(response, 'text', encoding = 'UTF-8'))
  if (is.null(data$value)) {
    stop("No data returned for indicator:", indicator_code)
  }
  
  # Determine the columns to select based on num_dims
  selected_columns <- c("SpatialDim", "TimeDim", "NumericValue")
  if (num_dims > 0){
    selected_columns <- c(selected_columns, paste0("Dim", seq_len(num_dims)))
  }
  
  # Process and reshape data
  df <- data$value %>%
    select(all_of(selected_columns)) %>%
    rename(
      country = SpatialDim,
      year = TimeDim,
      !!column_name := NumericValue
    )
  
  return(df)
}
```

### Fetch, Process, and Consolidate WHO Data for Multiple Indicators

#### Description

This code fetches and processes data for various health-related indicators from the World Health Organization (WHO) Global Health Observatory (GHO). The data spans key health metrics such as life expectancy, immunization coverage, health financing, alcohol consumption, birthweight estimates, anemia prevalence, and BMI statistics. Data sources for these indicators include:

- [Life Expectancy at Birth](https://www.who.int/data/gho/data/indicators/indicator-details/GHO/life-expectancy-at-birth-(years))
- [Immunization Coverage Estimates](https://www.who.int/data/gho/data/themes/topics/indicator-groups/indicator-group-details/GHO/immunization-coverage-estimates)
- [Prevalence of HIV](https://www.who.int/data/gho/data/themes/hiv-aids/data-on-the-size-of-the-hiv-aids-epidemic) [and Malaria](https://www.who.int/data/gho/data/themes/topics/indicator-groups/indicator-group-details/GHO/malaria-cases)
- [Health Financing Data](https://www.who.int/data/gho/data/themes/topics/health-financing)
- [Alcohol Consumption](https://www.who.int/data/gho/data/indicators/indicator-details/GHO/total-(recorded-unrecorded)-alcohol-per-capita-(15-)-consumption) [and Expenditure](https://www.who.int/data/gho/data/indicators/indicator-details/GHO/alcohol-expenditure-as-a-per-cent-of-total-household-expenditure)
- [Low Birthweight Estimates](https://www.who.int/data/gho/data/themes/topics/topic-details/GHO/low-birthweight-estimates)
- [Anaemia in Women and Children](https://www.who.int/data/gho/data/themes/topics/anaemia_in_women_and_children)
- [BMI Data for Children](https://www.who.int/data/gho/data/themes/topics/indicator-groups/indicator-group-details/GHO/bmi-among-children) [and Adults](https://www.who.int/data/gho/data/themes/topics/indicator-groups/indicator-group-details/GHO/bmi-among-adults)

#### Process

1. Iterating through a list of indicator metadata (name, code, and dimensions).
2. Using the fetch_numeric_data function to query the WHO API for each indicator.
3. Cleaning and reshaping data dynamically based on dimensionality requirements.
4. Combining all indicators into a single data frame using full joins on country and year.
5. Logging performance metrics for each indicator and the overall data-fetching process.
6. Exporting the consolidated data to a CSV file for further analysis.

```{r query-data-from-GHO}
# Initialize an empty list to store data for each indicator
numeric_data_list <- list()

# Define a list of indicators with associated metadata. Each entry contains:
# # Indicator Code
# # Descriptive Name
# # Number of Dimensions to Pivot (0 if no pivoting is needed)
indicators <- list(
  c('WHOSIS_000001', 'life_expectancy_at_birth', 1),
  
  c('WHS4_543', 'bcg_immunization_coverage', 0),
  c('WHS4_100', 'dtp_immunization_coverage', 0),
  c('WHS4_117', 'hepb3_immunization_coverage', 0),
  c('WHS4_129', 'hib3_immunization_coverage', 0),
  c('SDGHPVRECEIVED', 'hpv_immunization_coverage', 0),
  c('WHS8_110', 'mcv1_immunization_coverage', 0),
  c('MCV2', 'mcv2_immunization_coverage', 0),
  c('WHS4_128', 'pab_immunization_coverage', 0),
  c('PCV3', 'pcv3_immunization_coverage', 0),
  c('WHS4_544', 'pol3_immunization_coverage', 0),
  c('ROTAC', 'rotac_immunization_coverage', 0),
  
  c('MDG_0000000029', 'hiv_prevalence', 0),
  c('SDGHIV', 'new_hiv_per_1000', 0),
  c('MALARIA_EST_INCIDENCE', 'malaria_per_1000', 0),

  c('GHED_CHEGDP_SHA2011', 'che_%_of_gdp', 0),
  c('GHED_EXTCHE_SHA2011', 'ext_%_of_che', 0),
  c('GHED_GGHE-DCHE_SHA2011', 'gghe-d_%_of_che', 0),
  c('GHED_OOPSCHE_SHA2011', 'oop_%_of_che', 0),
  c('GHED_PVT-DCHE_SHA2011', 'pvt-d_%_of_che', 0),
  
  c('SA_0000001688', 'alc_consump_per_capita_liters', 1),
  c('SA_0000001476', 'alc_expenditure_%_of_household', 0),
  
  c('PRETERMBIRTH_RATE', 'preterm_birth_rate_per_100', 0),
  c('LBW_PREVALENCE', 'low_birth_weight_prevalence', 0),

  c('NUTRITION_ANAEMIA_PREGNANT_PREV', 'pregnant_women_anaemia_prevalence', 1),
  c('HEMOGLOBINLEVEL_PREGNANT_MEAN', 'pregnant_women_mean_hemoglobin', 0),
  c('NUTRITION_ANAEMIA_CHILDREN_PREV', 'children_anaemia_prevalence', 1),
  c('HEMOGLOBINLEVEL_CHILDREN_MEAN', 'children_mean_hemoglobin', 0),
  
  c('NCD_BMI_MINUS2C', 'thin_children_prevalence', 2),
  c('NCD_BMI_PLUS1C', 'overweight_children_prevalence', 2),
  c('NCD_BMI_PLUS2C', 'obese_children_prevalence', 2),
  
  c('NCD_BMI_18C', 'underweight_adults_prevalence', 1),
  c('NCD_BMI_25C', 'overweight_adults_prevalence', 1),
  c('NCD_BMI_30C', 'obese_adults_prevalence', 1)
)

# Record the start time of the entire data-fetching process
start_time <- Sys.time()

# Fetch and clean data for each indicator
for (indicator in indicators) {
  indicator_start_time <- Sys.time() # Track start time for this indicator
  
  df <- fetch_numeric_data(indicator[1], indicator[2], indicator[3])
  
  # Pivot data if the indicator has additional dimensions
  if (as.numeric(indicator[3]) > 0) {
    pivot_cols <- paste0("Dim", seq_len(as.numeric(indicator[3])))
    df <- df %>%
      pivot_wider(
        names_from = all_of(pivot_cols),
        values_from = indicator[2],
        names_sep = "_",
        names_prefix = paste0(indicator[2], "_")
      )
  }

  numeric_data_list[[indicator[2]]] <- df
  
  # Log completion with timestamp and duration
  indicator_end_time <- Sys.time()
  duration <- indicator_end_time - indicator_start_time
  cat(sprintf("Completed '%s' in %.2f seconds\n", indicator[2], as.numeric(duration, units = "secs")))
}

# Join all data frames on 'country' and 'year'
numeric_data <- reduce(numeric_data_list, function(x, y) {
  full_join(x, y, by = c("country", "year"))
})

# Log total duration of the data-fetching process
total_duration <- Sys.time() - start_time
cat(sprintf("Full query completed in %.2f seconds\n", as.numeric(total_duration, units = "secs")))
```

#### Export to a CSV

```{r export-data}
# Save the combined data to a CSV file
output_path <- "../../data/interim/WHO_data.csv"
write.csv(numeric_data, file = output_path, row.names = FALSE)
cat(sprintf("Data saved to '%s'\n", output_path))
```

